---
title: "Homework0"
author: "RyanLi"
format: html
editor: visual
execute: 
  echo: true
---

# Solutions of quadratic function

## Print out solution of quadratic function

a = 1, b= 3, c = 2

$$
y = x^2 + 3x + 2
$$

```{r}
find.roots = function(a, b, c) {
  discriminant = b^2 - 4 * a * c
  if (discriminant > 0) {
    c((-b - sqrt(discriminant))/(2 * a), (-b + sqrt(discriminant))/(2 * a))
  }
  else if (discriminant == 0) {
    -b / (2 * a)
  }
  else {
    NaN
  }
}
solutions = find.roots(1, 3, 2)
```

Solution: We first need to compute whether the discrimination is larger or equal to zero. $$ \delta = b^2 - 4ac = 1$$ Therefore, the quadratic function has two solutions. The solutions would be:

```{r}
solutions
```

# Quadratic function plot

```{r}

a = 1
b = 3
c = 2
f = function(x){
  a*x^2 + b*x + c
}
x = seq(-5, 5, length = 300)
plot(x, f(x), type = 'l')
abline(h = 0)
abline(v = 0)

points(x = solutions, y = rep(0, length(solutions)), # x and y coordinates of the x-intercepts
       pch = 18, cex = 2, col = 'pink')

text(x = solutions, y = rep(0, length(solutions)),
     labels = rep("x", length(solutions)),
     pos = 3, col = 'pink')
```

As we could see, there are two pink data points along the x-axis that represent the x-intercepts (solutions) of the quadratic function.
